% Topic 1.2.04 — Independence of Events.

\section{Independence of Events}
\label{sec:1.2.04-independence}

\begin{ticket}
Independence of events. Pairwise independence vs. joint (mutual) independence.
\end{ticket}

\subsection*{Theory}

Independence is the assumption that one piece of information carries
no probabilistic content about another — formally, that the joint
probability factors. The definition is purely multiplicative; no
``mechanism'' or causal interpretation is built in.

\begin{definition}[Two events]
\label{def:indep-two}
Events $A, B \in \mathcal{F}$ are \emph{independent} if
\[
  \Prob(A \cap B) \;=\; \Prob(A)\, \Prob(B).
\]
\end{definition}

\begin{proposition}[Conditional reformulation]
\label{prop:indep-conditional}
If $\Prob(B) > 0$, then $A, B$ are independent iff
$\Prob(A \mid B) = \Prob(A)$. Symmetrically, if $\Prob(A) > 0$, iff
$\Prob(B \mid A) = \Prob(B)$.
\end{proposition}
\proofof{indep-conditional}

\begin{proposition}[Independence is preserved under complementation]
\label{prop:indep-complements}
If $A, B$ are independent, so are the pairs $(A, B^c)$, $(A^c, B)$,
and $(A^c, B^c)$.
\end{proposition}
\proofof{indep-complements}

The extension from two events to a family of events admits two
natural interpretations, and they are \emph{not} equivalent.

\begin{definition}[Pairwise vs.\ mutual independence]
\label{def:indep-many}
Let $\{A_i\}_{i \in I}$ be a (finite or countable) family of events.
\begin{itemize}
  \item The family is \emph{pairwise independent} if
        $\Prob(A_i \cap A_j) = \Prob(A_i) \Prob(A_j)$ for every
        $i \neq j$ in~$I$.
  \item The family is \emph{(jointly or mutually) independent} if
        for every finite $J \subseteq I$,
        \[
          \Prob\Bigl(\bigcap_{i \in J} A_i\Bigr) \;=\; \prod_{i \in J} \Prob(A_i).
        \]
\end{itemize}
Mutual independence implies pairwise independence; the converse
fails for $|I| \geq 3$, as Bernstein's example shows
(\cref{ex:bernstein}).
\end{definition}

For mutual independence, one needs the factorisation to hold over
\emph{all} non-empty finite subsets, not just over the full
intersection. This is what makes the notion strong enough to
extend to operations on the events.

\begin{proposition}[Stability of mutual independence]
\label{prop:indep-stability}
Let $A_1, \dots, A_n$ be mutually independent. Then:
\begin{enumerate}[label=(\alph*)]
  \item replacing any $A_i$ by $A_i^c$ preserves mutual independence;
  \item if $1 \leq r < n$, the events $B = A_1 \cap \cdots \cap A_r$
        and $C = A_{r+1} \cap \cdots \cap A_n$ are independent (i.e.,
        a ``block'' independence statement).
\end{enumerate}
\end{proposition}
\proofof{indep-stability}

\medskip
\noindent\textbf{Independence of random variables.} Independence
extends to random variables via Borel preimages.

\begin{definition}
\label{def:indep-rv}
Random variables $X_1, \dots, X_n$ on $(\Omega, \mathcal{F}, \Prob)$
are \emph{(mutually) independent} if for every Borel sets
$B_1, \dots, B_n \in \mathcal{B}(\R)$,
\[
  \Prob(X_1 \in B_1, \dots, X_n \in B_n) \;=\; \prod_{i=1}^{n} \Prob(X_i \in B_i).
\]
Equivalently, in terms of joint and marginal CDFs
$F_{(X_1, \dots, X_n)}(x_1, \dots, x_n) = \prod_i F_{X_i}(x_i)$ for
all $x_i \in \R$.
\end{definition}

\begin{theorem}[Multiplicativity of expectation]
\label{thm:indep-multiplicative}
If $X, Y \in L^1$ are independent, then $XY \in L^1$ and
\[
  \E(XY) \;=\; \E X \cdot \E Y.
\]
Consequently, independent variables in $L^2$ are uncorrelated:
$\Cov(X, Y) = 0$. The converse is false in general
(\cref{ex:uncorrelated-dependent}).
\end{theorem}
\proofof{indep-multiplicative}

\begin{corollary}[Variance of a sum of independent variables]
\label{cor:var-sum-indep}
If $X_1, \dots, X_n \in L^2$ are mutually independent, then
\[
  \Var\Bigl(\sum_{i=1}^{n} X_i\Bigr) \;=\; \sum_{i=1}^{n} \Var X_i.
\]
\end{corollary}
\proofof{var-sum-indep}

\begin{remark}
Independence is the working assumption behind the limit theorems of
\cref{sec:1.2.05-limit-theorems}, the statistical-sampling models
of \cref{sec:1.2.10-chi2-f-student}, and the chain-of-conditionals
formalism of \cref{sec:1.2.09-markov-chains}, where it is replaced
by the weaker Markov property. For independence of $\sigma$-algebras
and the associated extension theorems, see \cite[Ch.~II,
\S~5]{Shiryaev1989}.
\end{remark}

\subsection*{Proofs}

\begin{proof}[\Cref{prop:indep-conditional}]
\proofanchor{indep-conditional}
$\Prob(A \mid B) = \Prob(A \cap B) / \Prob(B)$ by definition. Equality
to $\Prob(A)$ is therefore equivalent (under $\Prob(B) > 0$) to
$\Prob(A \cap B) = \Prob(A) \Prob(B)$, which is the independence
condition. The roles of $A$ and~$B$ are symmetric in the
factorisation form.
\end{proof}

\begin{proof}[\Cref{prop:indep-complements}]
\proofanchor{indep-complements}
$A = (A \cap B) \sqcup (A \cap B^c)$ is a disjoint decomposition, so
\[
  \Prob(A \cap B^c) = \Prob(A) - \Prob(A \cap B) = \Prob(A) - \Prob(A) \Prob(B) = \Prob(A) (1 - \Prob(B)) = \Prob(A) \Prob(B^c),
\]
using \cref{def:indep-two} and the complement rule
\cref{prop:prob-elementary}\,(c). Hence $A$ and $B^c$ are
independent. Apply this twice (first to $A$, $B$; then to
$B^c$, $A$) to deduce the other two pairs.
\end{proof}

\begin{proof}[\Cref{prop:indep-stability}]
\proofanchor{indep-stability}
\emph{(a)} It suffices to swap $A_n$ for $A_n^c$; iterating gives
the general claim. For any $J \subseteq \{1, \dots, n-1\}$, by
mutual independence applied to $J \cup \{n\}$ and to~$J$,
\[
  \Prob\Bigl(\bigcap_{i \in J} A_i \cap A_n^c\Bigr) = \Prob\Bigl(\bigcap_{i \in J} A_i\Bigr) - \Prob\Bigl(\bigcap_{i \in J} A_i \cap A_n\Bigr) = \prod_{i \in J} \Prob(A_i)\, (1 - \Prob(A_n)),
\]
which is exactly the factorisation for the family with $A_n$
replaced by $A_n^c$.

\emph{(b)} Expand using inclusion–exclusion-style arguments: by
mutual independence of $A_1, \dots, A_n$ applied to the full set
$\{1, \dots, n\}$,
\[
  \Prob(B \cap C) = \Prob\Bigl(\bigcap_{i=1}^{n} A_i\Bigr) = \prod_{i=1}^{n} \Prob(A_i) = \prod_{i=1}^{r} \Prob(A_i) \cdot \prod_{i=r+1}^{n} \Prob(A_i) = \Prob(B)\, \Prob(C),
\]
where the third equality used mutual independence on the prefix and
suffix subsets to compute $\Prob(B)$ and $\Prob(C)$ separately.
\end{proof}

\begin{proof}[\Cref{thm:indep-multiplicative}]
\proofanchor{indep-multiplicative}
Take first $X = \mathbf{1}_A$, $Y = \mathbf{1}_B$ for events
$A, B$. By independence of the $\sigma$-algebras
generated by $X$ and $Y$ (which for indicators is just
\cref{def:indep-two}), $\E(XY) = \E \mathbf{1}_{A \cap B} = \Prob(A \cap B) = \Prob(A) \Prob(B) = \E X \cdot \E Y$.
By linearity (\cref{prop:exp-properties}\,(a)), the identity extends
to simple functions
$X = \sum_{i} a_i \mathbf{1}_{A_i}$, $Y = \sum_{j} b_j \mathbf{1}_{B_j}$
with $\{A_i\}, \{B_j\}$ each a partition and the two partitions
``decoupled'' through independence of the corresponding
$\sigma$-algebras. The general $L^1$ case is obtained by approximating
$X, Y \geq 0$ by simple non-negative functions
$X_n \uparrow X, Y_n \uparrow Y$ and applying monotone convergence
(legitimate because each $X_n Y_n \leq XY$ pointwise, and
$\E X_n Y_n = \E X_n \cdot \E Y_n \to \E X \cdot \E Y$). Decompose
$X = X^+ - X^-$, $Y = Y^+ - Y^-$ to handle signs; full details
in \cite[Ch.~II, \S~6]{Shiryaev1989}.

The covariance corollary follows directly:
$\Cov(X, Y) = \E XY - \E X\, \E Y = 0$.
\end{proof}

\begin{proof}[\Cref{cor:var-sum-indep}]
\proofanchor{var-sum-indep}
By \cref{prop:cov-properties}\,(e),
$\Var(\sum X_i) = \sum \Var X_i + 2 \sum_{i < j} \Cov(X_i, X_j)$.
Pairs $(X_i, X_j)$ with $i \neq j$ are pairwise independent (the
projection of mutual independence to two indices), hence
uncorrelated by \cref{thm:indep-multiplicative}, so the off-diagonal
sum vanishes.
\end{proof}

\subsection*{Examples}

\begin{example}[Two coin tosses]
\label{ex:two-coins}
Toss a fair coin twice independently; let $A = \{\text{first is H}\}$
and $B = \{\text{second is H}\}$. Take $\Omega = \{H, T\}^2$ with the
classical measure. Then $\Prob(A) = \Prob(B) = 1/2$ and
$\Prob(A \cap B) = 1/4 = \Prob(A) \Prob(B)$, so $A$ and~$B$ are
independent — a sanity check that the model construction matches the
intuitive notion. Equivalently, $\Prob(A \mid B) = 1/2 = \Prob(A)$
(\cref{prop:indep-conditional}).
\end{example}

\begin{example}[Bernstein: pairwise but not mutually independent]
\label{ex:bernstein}
Take $\Omega = \{1, 2, 3, 4\}$ uniformly distributed,
$A = \{1, 2\}$, $B = \{1, 3\}$, $C = \{1, 4\}$. Each has probability
$1/2$, every pairwise intersection equals $\{1\}$ with probability
$1/4 = (1/2)(1/2)$, so the family is pairwise independent. But
$A \cap B \cap C = \{1\}$ with $\Prob(A \cap B \cap C) = 1/4 \neq 1/8 = (1/2)^3$.
The triple is therefore \emph{not} mutually independent — a
canonical illustration that the extra factorisations in
\cref{def:indep-many} cannot be reduced to pairwise checks.
\end{example}

\begin{example}[Number of heads, revisited]
\label{ex:binomial-indep}
With $S_n = X_1 + \cdots + X_n$ as in \cref{ex:binomial-moments},
the $X_i$ are mutually independent (by construction of independent
trials); \cref{cor:var-sum-indep} yields
$\Var S_n = \sum_i \Var X_i = n p(1 - p)$. The earlier derivation in
\cref{ex:binomial-moments} only invoked pairwise uncorrelatedness,
which is a strictly weaker hypothesis but happens to be enough for
the variance-of-sum formula.
\end{example}

\begin{example}[Uncorrelated does not imply independent]
\label{ex:uncorrelated-not-indep}
The pair $X \sim \mathrm{Uniform}\{-1, 0, 1\}$ and $Y = X^2$ from
\cref{ex:uncorrelated-dependent} are uncorrelated yet
deterministically related, so they are not independent: for instance,
$\Prob(X = 0, Y = 0) = \Prob(X = 0) = 1/3$, but
$\Prob(X = 0)\, \Prob(Y = 0) = (1/3)(1/3) = 1/9$.
\end{example}

\begin{problem}
\label{prob:exactly-k-out-of-n}
Roll a fair die $n$ times independently. Find the probability that the
number of sixes equals~$k$, and verify the resulting distribution sums
to~$1$ over $k = 0, 1, \dots, n$.
\end{problem}

\begin{proof}[Solution]
Let $X_i = \mathbf{1}_{\{\text{roll }i\text{ shows }6\}}$. Then
$X_1, \dots, X_n$ are mutually independent, each with parameter
$p = 1/6$. The total $S_n = \sum X_i$ counts sixes; the event
$\{S_n = k\}$ is the union over the $\binom{n}{k}$ choices of which
$k$ rolls are sixes, and on each such configuration the joint
probability factorises (mutual independence) as
$p^k (1-p)^{n-k}$. The disjoint union gives
\[
  \Prob(S_n = k) \;=\; \binom{n}{k} p^k (1 - p)^{n-k}, \qquad p = 1/6.
\]
By the binomial theorem,
$\sum_{k=0}^n \binom{n}{k} p^k (1-p)^{n-k} = (p + (1 - p))^n = 1$.
This is the binomial distribution; it returns in
\cref{sec:1.2.06-standard-distributions} as a standard discrete
family.
\end{proof}
